Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-04-05 pubs/tfp16/paper.tex
edits

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the US
government commits to a push for exascale computing ($10^{18}$ FLOPS in a
single system) within the next decade~\cite{Exascale-roadmap} a highly
respected member of the HPC community recently stated publicly that ``Fortran
is essential for exascale programming.''~\cite{Heroux16}\footnote{Various EU,
  Japanese, and international groups are also anticipating the realization
  exascale computing.~\cite{EUexascale,Exascale-org}} Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are reportedly in the same state.}  C++ has become more common
for new code starts, but their nature remains the same: huge codes with fair
to non-existent encapsulation of side effects, with multiple levels of
parallelism also poorly abstracted, especially at the thread (shared-memory)
level.  Thread-level parallel efficiency can be good when the parallel model
is bulk synchronous; tasking models are generally avoided unless abstracted
by, e.g., a C++ template library.

\subsection{A trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjunction
with the International Conference on Functional Programming, being now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that entire blocks be brought into scope.  It is almost ironic,
then, that of the mainstream high-performance scientific programming languages
(Fortran, C, and C++), Fortran~95 was the first to introduce the \texttt{pure}
function qualifier, and with enforcement by the compiler.  Otherwise it is
entirely up to the programmer to make functions composable and thread safe.
Here at LANL the use of pure functions (among other techniques for reducing
global data dependencies) is being used to good effect to clean up some legacy
code bases.

@@

Using Legion, all non-function-local variables, or \emph{regions}, must be
requested from the runtime system.  Regions may be shared among \emph{tasks}
(essentially C++ functions), and each task is prescribed access privileges
(read, write, read/write, etc.) and coherency requirements for each region to
which it will have access.  Other than registered access to regions, tasks
must not access non-constant global data.  In serial execution these access
requirements are notionally superfluous, but in parallel execution the runtime
can dynamically calculate the data dependency graph and relax the
programmatically defined serial ordering of task execution.  Very high
parallel efficiency has been demonstrated on extremely large computing
systems~\cite{Bauer14}.  Here at LANL the Legion system is being strongly
considered for a new large-scale C++ code effort.\footnote{It seems
  significant that the PI of the Legion project was previously a co-developer
  of a strict, arguably pure functional language
  implementation~\cite{AikenFL,FLreport89}, in collaboration with John Backus
  as a refinement of his FP language~\cite{Backus:1978}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practitioners who are developing an appreciation for pure functional
programming in Haskell.  Still, they see their needs as different than other
Haskell users: performance is paramount, vectors (``Efficient
Arrays''~\cite{vector}) are a (perhaps \emph{the}) primary data structure, and
non-strictness or laziness is more of a nuisance---even a serious
hindrance---than a convenience, much less ever algorithmically essential in
the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's general attribute
grammars~\cite{Johnsson87}, or other instances of \emph{tying the
  knot}~\cite{tying-the-knot}.

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get I/O and inter-process communication to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of execution, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece (whether
they've read it or not)~\cite{Hughes89}, they would also be happy to write in
a slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

Much anecdotal evidence suggests that many others have similar difficulties
with semantics that are lazy (or non-strict) by default, with the most telling
recent concrete evidence being the push for, and realization of, the
\texttt{Strict} and \texttt{StrictData} language extensions newly available in
GHC 8.0.  Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practitioners are by and large not computer
  scientists}, they are domain scientists (physicists, chemists, etc.) or
computational scientists specialized in numerical or numerically-oriented
algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
The concept of a space leak, for example, caused by other than a missing
explicit deallocation statement (C \emph{free}, etc.) is difficult to
understand and not something they particularly want to try to understand.
%Hence, we eschew explicit strictness annotations for varying uniform degrees
%of strictness.

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model typically implemented with the Message Passing Interface
(MPI) \cite{MPI}.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  A function in a pure functional language
could be regarded as the essence of the concept of a task.  All of these
systems are of course in languages with strict function-call semantics.

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various points in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can be (manually) restarted from the
last stored state.  Most implementations are explicit, though there has been
some success with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating GHC language, type
system, and runtime features, or significant code optimization, would simply
divert us from our central goals.\footnote{We estimate that we will have, to
  within an order of magnitude, one person-hour of total available development
  time for each person-year that GHC currently represents, or put another way,
  one second for each hour, respectively.}  For another, it means that we are
not undertaking significant morphing of GHC itself: because this is in large
part a student project it must be kept as simple as reasonably possible in
various respects.  Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}

For feasibility, we restrict ourselves to shared-memory implicit parallelism.
The requirement for tail calling is not primarily for efficiency, rather it
is for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, makes this an interesting
exercise that we will discuss at some length.

@@

In a strict semantics one could devise various evaluation orders, but as far
as distinguishing termination properties these can be distilled down to two
reasonable possibilities, namely whether on construction of a \emph{PAP} the
(new) arguments are forced or not.  In other words, the question is whether
application is strict, or actual function call is strict.  These are in fact
distinguishable because (in Haskell) the first argument to \texttt{seq} might
evaluate to a \emph{PAP}.

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict; function strict, in which only
those arguments needed by a function call are evaluated; and application
strict, in which all arguments of an application are evaluated.  The other
dimension we vary is whether or not constructors are strict.  Other dimensions
are possible, for example whether \texttt{let} bindings are strict, but we
restrict ourselves to function/application and constructor strictness.  The
STG machinery makes the choice of evaluation strategy easy to change, both in
terms of function application and building a constructor in the heap.

\subsection{STG to C}
The back-end is written in Haskell and generates C code;
the runtime system is written in C/C++. This allows us to leverage the extensive
optimizers in modern C compilers such as gcc and Clang/LLVM. The system fairly 
faithfully implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which, of course, may be different than those for non-strict
semantics.  Sharing is preserved.

\subsubsection{Proper tail calls.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel 
execution they identified the need for the ability to examine and unwind the 
stack.  This requires that use of the C stack be highly constrained if we are 
to avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
non-portable code to directly manipulate the C stack, complicating an
already non-trivial implementation effort.  We therefore must limit the
use of the C stack to calls to the runtime, in other words, when executing
code generated from the user program (and not in a runtime call), there
must be no growing of (pushing return addresses or data onto) the C stack.

First we establish that this is possible in principle.  By maintaining our own
data and control (continuation) stack all generated C functions can take no
explicit arguments and have return type \emph{void}. If we arrange that
generated C code has only calls (other than runtime calls, which do not call
back to user-program code) in tail call positions, that is, immediately before
a return in the control flow graph, then at non-zero optimization levels both
gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When
using gcc we can also use a fixed set of machine registers for
passing/returning values, much like GHC.

\subsubsection{A taste of the continuation style.}
Now to implementation.  To clearly distinguish between a notional C call (a
syntactic concept) and a machine-level call (pushing a return address onto the
native C stack) we will refer to the latter as \emph{call-with-return}, and to
a machine-level tail call as \emph{jump}.

@@

Our solution to this entails passing \emph{stgApply} the object to be applied
in \emph{CV}, together with an extra argument---call it
\emph{index}---indicating which of the object to be applied and its boxed
arguments have been already been evaluated (previously by \emph{stgApply}).
If the object to be applied, or some argument (because of the strictness
level), remains to be evaluated, it is jumped to, with return as ever through
the continuation stack.  This is actually fairly efficient because we can
arrange that the stack frame for \emph{stgApply} is reused, updating only
\emph{index}, so except for the indirect jumps on return this is essentially a
loop.  For the cases of arguments in excess of the arity of the object being
applied, in effect \emph{stgApply}'s argument frame is popped and two
continuations are pushed, the first one to jump to \emph{stgApply} with the
excess arguments, the next an argument frame for the object to applied, then
the code for the object to be applied is jumped to.

>>>>

\section{Background and Motivation}
The practice of so-called high-performance scientific computing (HPC) is,
overall, highly conservative with respect to change.  Even as the USA and
other governments commit to a push for exascale computing ($10^{18}$ FLOPS in
a single system) within the next
decade~\cite{Exascale-roadmap,EUexascale,Exascale-org} a highly respected
member of the HPC community recently stated publicly that ``Fortran is
essential for exascale programming.''~\cite{Heroux16} Existing US government
laboratory HPC codes run to hundreds of thousands of lines of Fortran each and
complete rewrites are infeasible.\footnote{The codes of our lab's British
  counterpart are similar.}  C++ has become more common for new code starts,
but their nature remains the same: huge codes with fair to non-existent
encapsulation of side effects, with multiple levels of parallelism sometimes
poorly abstracted, especially at the thread (shared-memory) and computational
accelerator (e.g., GPU) levels.  Thread-level parallel efficiency can be good
when the parallel model is bulk synchronous; tasking models are generally
avoided unless abstracted by, e.g., a C++ template or runtime library, because
of the programming complexity.

\subsection{A trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new as evidenced
by the Workshop on Functional High Performance Computing, held in conjunction
with the International Conference on Functional Programming, being now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

\subsubsection{Rely on programmer discipline.}
Legacy Fortran scientific codes likely represent worst practices in state
encapsulation, with most data global and accessible to all; indeed, Fortran
was designed to make this easy via \texttt{COMMON} blocks---aggregates of
global variables---which further obfuscate program meaning by allowing
arbitrary naming of a block's variables at a subroutine/function granularity,
and requiring that entire blocks be brought into scope.  It is almost ironic,
then, that of the mainstream high-performance scientific programming languages
(Fortran, C, and C++), Fortran~95 was the first to introduce the \texttt{pure}
function qualifier, and with enforcement by the compiler.  Otherwise it is
entirely up to the programmer to make functions composable and thread safe.
%Here at LANL the use of pure functions (among other techniques for reducing
%global data dependencies) is being used to good effect to clean up some legacy
%code bases.

@@

Using Legion, all non-function-local variables, or \emph{regions}, must be
requested from the runtime system.  Regions may be shared among \emph{tasks}
(essentially C++ functions), and each task is prescribed access privileges
(read, write, read/write, etc.) and coherency requirements for each region to
which it will have access.  Other than registered access to regions, tasks
must not access non-constant global data.  Thus the inputs and outputs of
tasks are exactly known to the runtime system.  In serial execution these access
requirements are notionally superfluous, but in parallel execution the runtime
can dynamically calculate the data dependency graph and relax the
programmatically defined serial ordering of task execution.  Very high
parallel efficiency has been demonstrated on extremely large computing
systems~\cite{Bauer14}.\footnote{It seems
  significant that the PI of the Legion project was previously a co-developer
  of a strict, arguably pure functional language
  implementation~\cite{AikenFL,FLreport89}, in collaboration with John Backus
  as a refinement of his FP language~\cite{Backus:1978}.}

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be of some import, there is now a small number of laboratory HPC
practitioners who are developing an appreciation for pure functional
programming in Haskell.  Still, they see their needs as different than other
Haskell users:  performance is paramount, vectors (``Efficient
Arrays''~\cite{vector}) are a (perhaps \emph{the}) primary data structure, and
non-strictness or laziness is more of a nuisance---even a serious
hindrance---than a convenience, much less ever algorithmically essential in
the sense of Bird's \emph{repmin}~\cite{Bird84}, Johnsson's general attribute
grammars~\cite{Johnsson87}, or other instances of \emph{tying the
  knot}~\cite{tying-the-knot}.

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get I/O and inter-process communication to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of evaluation, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece (whether
they've read it or not)~\cite{Hughes89}, they would also be happy to write in
a slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

Much anecdotal evidence suggests that many others have similar difficulties
with semantics that are lazy (or non-strict) by default, with the most telling
recent concrete evidence being the push for, and realization of, the
\texttt{Strict} and \texttt{StrictData} language extensions newly available in
GHC 8.0.  Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practitioners are generally not computer
  scientists}, they are domain scientists (physicists, chemists, etc.) or
computational scientists specialized in numerical or numerically-oriented
algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
The concept of a space leak, for example, caused by other than a missing
explicit deallocation statement (C \emph{free}, etc.) is difficult to
understand and not something they particularly want to try to understand.
%Hence, we eschew explicit strictness annotations for varying uniform degrees
%of strictness.

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model usually implemented with the Message Passing Interface
(MPI)~\cite{MPI}\@.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  A function in a pure functional language
could be regarded as the essence of the concept of a task.  All of these
systems are of course in languages with strict function-call semantics.

\subsection{Another trend: transparent fault tolerance}
In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various points in time sufficient program state is dumped to backing store
such that the program, in case of a fault, can be (manually) restarted from
the last stored state.  Most implementations are explicit, though there has
been some success with transparent (invisible to the programmer) systems.
Wiring checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of the application
because it must be designed such that essential state can be readily captured.
Furthermore, there is a concern that the scalability of checkpoint/restart is
reaching its end.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating GHC language, type
system, and runtime features, or significant code optimization, would simply
divert us from our central goals.\footnote{We estimate that we will have, to
  within an order of magnitude, one person-hour of total available development
  time for each person-year that GHC currently represents, or put another way,
  one second for each hour, respectively.}  For another, it means that we are
not undertaking significant morphing of GHC itself: because this is in large
part a student project it must be kept as simple as reasonably possible in
various respects.  Our initial requirements are that
%
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}
%
For feasibility, we restrict ourselves to shared-memory implicit parallelism.
The requirement for tail calling is not primarily for efficiency, rather it
is for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, makes this an interesting
exercise that we will discuss at some length.

@@

In a strict semantics one could devise various evaluation orders, but as far
as distinguishing termination properties it seems these can be distilled down
to two reasonable possibilities, namely whether on construction of a
\emph{PAP} the (new) arguments are forced or not.  In other words, the
question is whether application is strict, or actual function call is strict.
These are in fact distinguishable because (in Haskell) the first argument to
\texttt{seq} might evaluate to a \emph{PAP}.

This then gives three reasonable operational modes that are strictly ordered
in terms of termination properties: non-strict; function strict, in which only
those arguments needed by a function call are evaluated; and application
strict, in which all arguments of an application are evaluated.  The other
dimension we vary is whether or not constructors are strict.  Other dimensions
are possible, for example whether \texttt{let} bindings are strict, but we
restrict ourselves to function/application and constructor strictness.  The
STG machinery makes the choice of evaluation strategy easy to change.
%, both in terms of function application and building a constructor in the
%heap.

\subsection{STG to C}
The back-end is written in Haskell and generates C code; the runtime system is
written in C/C++. This gives us the extensive optimization performed by modern
C compilers such as gcc and Clang/LLVM. The system fairly faithfully
implements the STG machine as described in \emph{EA} and dynamically
illustrated by Pope's Ministg~\cite{ministg} with the exception of the state
transition rules which may be different than those for non-strict semantics.
Sharing is preserved.

\subsubsection{Proper tail calls in C}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel
execution they identified the need for the ability to examine and unwind the
stack in quashing concurrent evaluation of the same \emph{THUNK}.  This
requires that use of the C stack be highly constrained if we are to avoid
complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or non-portable
code to directly manipulate the C stack, complicating an already non-trivial
implementation effort.  We therefore must limit the use of the C stack to
calls to the runtime, in other words, when executing code generated from the
user program (and not in a runtime call), there must be no growing of (pushing
return addresses or data onto) the C stack.
%
To clearly distinguish between a notional C call (a syntactic concept) and a
machine-level call (pushing a return address onto the native C stack) we will
refer to the latter as \emph{call-with-return}, and to a machine-level tail
call as \emph{jump}.

First we establish that this is possible in principle.  By maintaining our own
data and control (continuation) stack all generated C functions can take no
explicit arguments and have return type \emph{void}. If we arrange that
generated C code has only calls (other than runtime calls, which do not call
back to user-program code) in tail call positions, that is, immediately before
a return in the control flow graph, then at non-zero optimization levels both
gcc and Clang/LLVM will generate jumps instead of calls-with-return for C
calls.  When using gcc we can also use a fixed set of machine registers for
passing/returning values, much like GHC using the LLVM back-end with its
custom GHC calling convention.

The STG abstract machine is implemented using a continuation stack.
Continuations contain a code pointer, optionally some data, and layout
information for the benefit of the garbage collector.  For example, the
argument frame for a function call is one type of continuation.  When the
evaluation of a C function (corresponding to some STG code) is completed its
return is by jumping through the code pointer of the continuation on the top
of the stack (``returning through the continuation stack'').  In the case of
an argument frame for a function, the associated code simply pops the
continuation off the stack and returns through the next continuation on the
stack.

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  Here the object to be applied is first evaluated with
call-with-return.  When a resulting \emph{FUN} is applied to more than its
arity number of arguments, it is called-with-return with its arity number of
arguments, and the resulting value applied to the remaining arguments.  When
considering more strict semantics call-with-return is even more convenient
because arguments can be evaluated in a tight loop.  Unfortunately these uses
of call-with-return can result in the C stack growing unboundedly.  Our
solution uses a single C \emph{stgApply} function with no calls-with-return.
This is straightforwardly realizable because the stack is explicit, and with
the C compiler call optimization, the continuation style provides a proper
\emph{goto} mechanism where the \emph{code labels}---entities that name blocks
of code, can be treated as first class (e.g.\ stored), and serve as targets of
a jump---are C function addresses (c.f.\ \S 6.1~\cite{PJ-stockhardware}).

\begin{comment}
\subsubsection{A taste of the continuation style.}
Now to implementation.  To clearly distinguish between a notional C call (a
syntactic concept) and a machine-level call (pushing a return address onto the
native C stack) we will refer to the latter as \emph{call-with-return}, and to
a machine-level tail call as \emph{jump}.

@@

Our solution to this entails passing \emph{stgApply} the object to be applied
in \emph{CV}, together with an extra argument---call it
\emph{index}---indicating which of the object to be applied and its boxed
arguments have been already been evaluated (previously by \emph{stgApply}).
If the object to be applied, or some argument (because of the strictness
level), remains to be evaluated, it is jumped to, with return as ever through
the continuation stack.  This is actually fairly efficient because we can
arrange that the stack frame for \emph{stgApply} is reused, updating only
\emph{index}, so except for the indirect jumps on return this is essentially a
loop.  For the cases of arguments in excess of the arity of the object being
applied, in effect \emph{stgApply}'s argument frame is popped and two
continuations are pushed, the first one to jump to \emph{stgApply} with the
excess arguments, the next an argument frame for the object to applied, then
the code for the object to be applied is jumped to.
\end{comment}
